Equations & R Code

### 2016-12-28 manuscript/coexistence-stability_ms.tex
updates fig 1 background to white
updates ms with PBA edits

MacArthur (1955), Elton (1958), and even Darwin (Turnbull et al. 2013)
recognized the potential for compensatory dynamics among species to
stabilize ecosystem functioning in fluctuating environments. This idea
underlies the ``insurance hypothesis'' (Yachi and Loreau 1999), which
suggests stability increases with diversity because species respond
dissimilarly to environmental conditions, broadening the range of
conditions under which the community maintains function (Loreau 2010). A
variety of theoretical models all predict a positive relationship
between species richness and ecosystem stability (Lehman and Tilman
2000, Ives and Hughes 2002, Loreau and {{de Mazancourt}} 2013), and
experimental tests tend to support such a prediction (Tilman et al.
2006, Hector et al. 2010).

However, the ability of biodiversity-ecosystem functioning (BEF)
experiments to accurately represent real-world dynamics is debated
(Wardle 2016, Eisenhauer et al. 2016). Much of the debate centers around
the fact that BEF experimental protocols do not allow species gains to
offset species losses. Theoretical work on diversity-stability
relationships typically suffers from the same limitation: it recognizes
the role of environmental variability in driving population fluctuations
which destabilize ecosystems, but ignores the potential for
environmental variability to promote species richness and thereby help
stabilize ecosystems (Loreau 2010, but see Chesson et al. 2001).

Fluctuating environmental conditions are an important ingredient for
stable species coexistence, both in theoretical models (Chesson 2000a,
Chesson et al. 2004) and in natural communities (C{á}ceres 1997,
Descamps-Julien and Gonzalez 2005, Adler et al. 2006, Angert et al.
2009). Such ``fluctuation-dependent'' coexistence requires that species
have unique environmental responses and that environmental conditions
vary so that each species in a community experiences favorable and
unfavorable conditions, which prevents competitive exclusion (Chesson
2000a). When coexistence is maintained by a fluctuation-dependent
mechanism, an increase in environmental variability might lead to an
increase in species richness and, consequently, an increase in ecosystem
stability. However, increasing environmental variability may also
decrease ecosystem stability by increasing the fluctuations of
individual species, regardless of species richness. These countervailing
effects of environmental variability present an interesting paradox:
increasing variability should decrease ecosystem stability, but the
potential for an increase in richness might offset the decrease in
stability. Such a paradox complicates predictions about how ecosystems
will respond to predicted departures from the historical ranges of
environmental variability.

The opposing effects of environmental variability on stability might
explain the mixed results from empirical studies on the
diversity-stability relationship. Observational tests of the
diversity-stability relationship, which require sampling across natural
diversity gradients, have yielded positive (Hautier et al. 2014),
neutral (Valone and Hoffman 2003, Cusson et al. 2015), and negative
(Sasaki and Lauenroth 2011) relationships. In a meta-analysis of
diversity-stability relationships based on observational studies in
terrestrial ecosystems, Jiang and Pu (2009) found no significant
evidence for an effect of species richness on ecosystem stability. The
idiosyncratic results of these empirical studies contrast with the
consistent conclusions from theoretical work which ignores feedbacks
between variability and richness.

The gap between theoretical expectations and empirical results of
richness-stability relationships reflects the divergence of theory
developed to explain species coexistence and theory developed to explain
diversity and stability. One reason these two disciplines have diverged
is because they have focused on different questions. Diversity-stability
studies typically ask how ecosystem variability responds to different
levels of species richness at a given level of environmental variability
(reviewed in Kinzig et al. 2001, Loreau 2010), whereas coexistence
studies ask how species richness responds to different levels of
environmental variability (Chesson and Warner 1981).

To reconcile these two perspectives, we extend theory on the
relationship between species richness and ecosystem stability to cases
in which species coexistence explicitly depends on environmental
fluctuations and species-specific responses to environmental conditions.
We focus on the temporal storage effect and relative nonlinearity using
a general consumer-resource model. We use the model to investigate three
questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Does the diversity-stability relationship remain positive when species
  coexistence is fluctuation-dependent?
\item
  When species coexistence is fluctuation-dependent, how does increasing
  environmental variability impact ecosystem stability?
\item
  Do our answers to the previous two questions depend on the specific
  fluctuation-dependent coexistence mechanism (i.e., storage effect
  vs.~relative nonlinearity)?
\end{enumerate}

@@

During a growing season, consumer-resource dynamics are modeled as two
differential equations:

\begin{align}
\frac{\text{d}N_{i}}{\text{d}t} &= \epsilon_if_{i}(R)N_{i}, \quad t \ne \tau_k\\
\frac{\text{d}R}{\text{d}t} &= - \sum\limits_{i}f_{i}(R)N_{i}, \quad t \ne \tau_k
\end{align}

@@

Along with resource uptake, consumer population growth depends on the
production of dormant biomass (\(D\)), the activation of dormant biomass
to live biomass (\(D \rightarrow N\)), and the survival of living
biomass from one year to the next. The biomass of each species' states
at the start of a growing season are equal to

\begin{align}
  D_{i}(\tau_k^+) &= (1-\gamma_{i,\tau_k})[\alpha_i N_{i}(\tau_k) + D_{i}(\tau_k)](1-\eta_i) \\
  N_{i}(\tau_k^+) &= (1-\alpha_i)N_{i}(\tau_k) + \gamma_{i,t}[\alpha_i N_{i}(\tau_k) + D_{i}(\tau_k)] (1-\eta_i),
\end{align}

@@

For the storage effect to operate, we need species-specific responses to
environmental variability, density-dependent covariance between
environmental conditions and competition (\emph{EC} covariance), and
subadditive population growth (Chesson 1994, 2000b). Regardless of the
mechanism, long-term coexistence is possible when all species can
increase when rare. In the storage effect, rare species increase by
escaping the effects of \emph{EC} covariance. This happens because
common species will experience greater than average competition
(\emph{C}) in good environment (\emph{E}) years because common species
cannot avoid intraspecific competition. Rare species do not have this
problem and can increase rapidly in a good \emph{E} year. \emph{EC}
covariance is included in our model because dormant-to-live transition
rates (\(\gamma\)) are species-specific and vary through time. In a high
\(\gamma\) year for a common species, resource uptake will be above
average, while in a high \(\gamma\) year for a rare species, resource
uptake will be below average.

Subadditive population growth buffers populations against large
population decreases in unfavorable years. It is included in our model
through a dormant stage with very low death rates, which limits large
population declines. In combination, subadditive population growth
limits population declines in bad \emph{E} years, and \emph{EC}
covariance ensures species can increase rapidly when rare but suffer
when common.

We generated sequences of (un)correlated dormant-to-live state
transition rates (\(\gamma\)) for each species by drawing from
multivariate normal distributions with mean 0 and a variance-covariance
matrix (\(\Sigma(\gamma)\)) of

\begin{align}
\Sigma(\gamma) = 
\begin{bmatrix}
1 & \rho_{1,2} & \rho_{1,3} & \rho_{1,4} \\
\rho_{2,1} & 1 & \rho_{2,3} & \rho_{2,4} \\
\rho_{3,1} & \rho_{3,2} & 1  & \rho_{3,4} \\
\rho_{4,1} & \rho_{4,2} & \rho_{4,3} & 1  \\
\end{bmatrix}
\sigma_{E}^2
\end{align}

\noindent where \(\sigma^2_{E}\) is the variance of the environmental
cue and \(\rho_{i,j}\) is the correlation between species \emph{i}'s and
species \emph{j}'s transition rates. \(\rho\) must be less than 1 for
stable coexistence, and in all simulations we placed that constraint
that all \(\rho_{i,j}\)'s are equal. In a two-species community, the
inferior competitor has the strongest potential to persist when
\(\rho=-1\) (perfectly uncorrelated transition rates). However, in a
four-species community the minimum possible correlation among species is
-1/3 given our constraints that all \(\rho\)'s are equal and that
\(\Sigma(\gamma)\) must be positive-definite. We used the \texttt{R}
function \texttt{mvrnorm} to generate sequences of (un)correlated
variates \emph{E} that we converted to germination rates in the 0-1
range: \(\gamma = e^E / \left(1 + e^E \right)\).

@@

For example, in Fig. 1C we show two species' resource uptake curves that
are relatively nonlinear. Species B has the lowest \(R^*\) and would
competitively exclude species A in the absence of environmental
fluctuations. But, fluctuating resource supplies can benefit species A
because it can take advantage of relatively high resource levels due its
higher saturation point. Stable coexistence is only possible, however,
if each species creates a disadvantage for itself when abundant. This
occurs in our model because when a resource conservative species (e.g.,
species B in Fig. 1C) is abundant, resource levels will remain high for
a longer period of time because its draw down of resources saturates.
Likewise, when a resource acquisitive species (e.g., species A in Fig.
1C) is abundant, it quickly draws down resources to levels that favor
resource conservative species. Such reciprocity ensures species can
increase when rare and stabilizes coexistence (Armstrong and McGehee
1980, Chesson 2000a, Chesson et al. 2004).

@@

To explore how fluctuation-dependent coexistence can affect the
diversity-stability relationship, we simulated the model with four
species under two scenarios for each coexistence mechanism. First, we
allowed the variance of the environment to determine how many species
can coexist, akin to a community assembly experiment with a species pool
of four species. To do this, we simulated communities with all species
initially present across a gradient of annual resource variability (for
relative nonlinearity) or environmental cue variability (for the storage
effect). Second, we chose parameter values that allowed coexistence of
all four species and then performed species removals, akin to a
biodiversity-ecosystem function experiment. The two simulation
experiments correspond to (i) sampling ecosystem function across a
natural gradient of species richness and (ii) sampling ecosystem
function across diversity treatments within a site.

To understand how increasing environmental variability will impact
ecosystem stability when coexistence is fluctuation-dependent, we
simulated the model over a range of species pool sizes and environmental
cue or resource variability. For each size of species pool (1, 2, 3, or
4 species), we simulated the model at 15 evenly-spaced levels of
environmental cue (range = 0.1,2) for the storage effect and 25
evenly-spaced levels of resource variability (range = 0.1,1.4) for
relative nonlinearity. We also explored the influence of asymmetries in
species' competitive abilities and correlations in species'
environmental responses within the storage effect model.

Under relative nonlinearity, species' resource response curves (Fig.
S1-1) reflect traits that determine the variability each species' growth
through time. ``Stable'' species have conservative resource uptake
curves, whereas ``unstable'' species have highly nonlinear resource
uptake curves with a large difference in growth between low and high
resource availability. Therefore, we ran two sets of simulations for
relative nonlinearity: one where the species pools increased from stable
to unstable species and vice versa. For example, if species A is the
most stable species and species D is the least stable we ran simulations
where the species pool increased from one to four species as A then B
then C then D. We then ran simulations with that order reversed.

All simulations were run for 5,000 seasons with 100-day growing seasons.
We averaged biomass over the growing season, and those yearly values
were used to calculate total community biomass in each year. After
discarding an initial 500 seasons to reduce transient effects on our
results, we calculated the coefficient of variation (\emph{CV}) of
summed species biomass through time, which represents ecosystem
variability, the inverse of ecosystem stability. We calculated species
richness as the number of species whose average biomass was greater than
1 over the course of the simulation. Parameter values for specific
results are given in figure captions. Within-season dynamics were solved
given initial conditions using the package \texttt{deSolve} (Soetaert et
al. 2010) in \texttt{R} (Team 2013). \texttt{R} code for our model
function is in the Supporting Information Appendix S1. All model code
has been deposited on Figshare (\emph{link}) and is available on GitHub
at \url{http://github.com/atredennick/Coexistence-Stability}.

@@

When we allowed the variance of the environment to determine which of
four initial species coexist, we found a positive relationship between
richness and variability (Fig. 2A,C). This was true for the storage
effect, where coexistence is maintained by fluctuating transition rates
(\(\gamma\)), and for relative nonlinearity, where coexistence is
maintained by annual resource pulses. The relationship is driven by the
fact that, under both coexistence mechanisms, the strength of species
coexistence increased with environmental variability (Fig. S1-2).
Therefore, more variable conditions promoted species richness, creating
a positive relationship between diversity and variability.

When we performed species removals but held environmental variability at
given levels, we found a negative diversity-variability relationship
(Fig. 2B,D). Scatter around the relationship was small for the storage
effect because all species have similar temporal variances. Therefore,
regardless of species identity, having more species will always
stabilize ecosystem functioning through portfolio effects. On the
contrary, scatter around the relationship was larger for relative
nonlinearity (Fig. 2D) because each species has unique temporal variance
depending on its resource uptake curve. Depending on the specific
combination of species, two-species communities were sometimes less
variable than three-species communities.

To understand how much species additions can stabilize ecosystem
functioning, we simulated our model over a range of environmental
variance and species pool sizes. Realized species richness increased
with environmental variability and, in some cases, increasing
variability completely offset the effect of increasing environmental
variability on stability. More species rich communities were less
variable on average and increased in ecosystem \emph{CV} at a slower
rate than communities with fewer species (e.g., lower slopes in log-log
space; Fig. S1-3).

The dampening effect of fluctuation-dependent coexistence on increasing
environmental variability depends on the specific traits (parameter
values) of the species in the regional pool. Moderate competition makes
it more difficult for new species to enter the local community, but once
they do enter, variability is similar (Fig. 3; compare top and bottom
panels). Moderate competition also increases the rate at which ecosystem
\emph{CV} increases with environmental variance (Fig. S1-3). This is
because the abundance of inferior competitors is reduced and do not
influence ecosystem \emph{CV} as much as when competition is low. The
correlation of species' environmental responses also mediates the
relationship between environmental variance, species richness, and
ecosystem \emph{CV}: lower correlations make it easier for new species
to enter the community and unique environmental responses are always
stabilizing (Fig. 3).

In communities where species coexist via relative nonlinearity, whether
or not the direct impact of environmental variability on ecosystem
variability is tempered by species additions depends on the species
traits of immigrating species. When additional species, which immigrate
from the regional pool, are less intrinsically stable than the resident
species, ecosystem variability increases at a constant rate even as
species are added (Fig. 4A). On the contrary, if more stable species are
added, species additions buffer the ecosystem from increasing
environmental variability (Fig. 4B). When just the most unstable
(resource acquisitive) species is present, ecosystem variability
initially declines as resource variability increases (Fig. 4B). This is
because fluctuations between very low and very high biomass become less
frequent as the species' resource uptake is more consistently high. The
stability of individual species in our relative nonlinearity model is
determined by their respective resource response curves (Fig. S1-1).
Under relative nonlinearity, we find that the buffering effect of
species additions depends on species traits, and the order in which
species enter the local community. Indeed, if all species in the
regional pool are less stable than the resident species, then no
stabilization occurs as species are added (Fig. 4A).

@@

Theory developed for biodiversity-ecosystem function experiments
emphasizes that increases in species richness should reduce ecosystem
variability. Consistent with theoretical expectations from models in
which species coexistence is maintained by fluctuation-independent
mechanisms, we produced the typical negative diversity-variability
relationship (Fig. 2B,D) from a model with fluctuation-dependent species
coexistence (also see Chesson et al. 2001). Likewise, our results from
the species removal simulations are consistent with results from
biodiversity-ecosystem functioning experiments showing a negative
relationship between species richness and ecosystem variability. This is
encouraging because species almost certainly coexist by some combination
of fluctuation-independent (e.g., resource partitioning) and
fluctuation-dependent mechanisms. By extending theory to communities
where species richness is explicitly maintained by temporal variability,
we have gained confidence that experimental findings are generalizable
to many communities. In other words, in local settings where
environmental variability is relatively homogeneous, reductions in the
number of species will reduce the stability of ecosystem functioning,
regardless of how coexistence is maintained.

When we allowed a gradient of environmental variability to determine
species coexistence, we discovered a positive relationship between
species richness and ecosystem variability (Fig. 2A,C). While surprising
when viewed through the lens of previous theory and experimental
findings, such a relationship is a direct consequence of how diversity
can be maintained in fluctuating environments. The storage effect and
relative nonlinearity both require environmental fluctuations to allow
niche differentiation between species pairs (Chesson 2000a). Therefore,
species coexistence gains strength, for both mechanisms, as the
environment becomes more variable (Fig. S1-2), but the increase in
environmental variability also increases ecosystem variability.

Our results may explain why deviations from the negative
diversity-variability relationship often come from observational studies
(Jiang and Pu 2009). Observational studies must rely on natural
diversity gradients, and if species richness depends on environmental
variability, it is entirely possible to observe positive
diversity-variability relationships. For example, Sasaki and Lauenroth
(2011) found a negative relationship between species richness and the
temporal stability of plant abundance (a positive diversity-variability
relationship) in a semi-arid grassland. Their data came from a six sites
that were 6 km apart. While Sasaki and Lauenroth explained their results
in terms of dominant species' effects, it is also possible that each
site experienced sufficiently different levels of environmental
variability to influence species coexistence. DeClerck et al. (2006)
also found a positive diversity-variability when sampling conifer
richness and the variability of productivity across a large spatial
gradient in the Sierra Nevada.

While our modeling results show that fluctuation-dependent coexistence
can create positive diversity-variability relationships, whether such
trends are detected will depend on the particular traits of the species
in the community and the relative influence of fluctuation-dependent and
fluctuation-independent coexistence mechanisms, which are not mutually
exclusive. Thus, our results may also help explain observational studies
where no relationship between diversity and variability is detected. For
example, Cusson et al. (2015) found no relationship between species
richness and variability of abundances in several marine macro-benthic
ecosystems. Many of their focal ecosystems were from highly variable
intertidal environments. If coexistence was at least in part determined
by environmental fluctuations, then the confounding effect of
variability and species richness could compensate any direct effect of
species richness on variability. Previous theoretical work showed how
environmental variation can mask the effect of species diversity on
ecosystem productivity when sampling across sites (Loreau 1998). Our
mechanistic model extends that conclusion to ecosystem stability.

Whether coexistence is fluctuation-independent or fluctuation-dependent
becomes especially important when we consider how ecosystem stability
responds to increasing environmental variability. In the
fluctuation-independent case, species richness is essentially fixed
because the species' inequalities that determine coexistence (niche and
fitness differences) are not linked to environmental variability.
Therefore, increasing environmental variability will always increase
ecosystem variability by increasing the fluctuations of individual
species' abundances.

>>>>

MacArthur (1955), Elton (1958), and even Darwin (Turnbull et al. 2013)
recognized the potential for compensatory dynamics among species to
stabilize ecosystem functioning in fluctuating environments. This idea
underlies the ``insurance hypothesis'' (Yachi and Loreau 1999), which
suggests ecosystem variability decreases with diversity because species
respond dissimilarly to environmental variation, broadening the range of
conditions under which the community maintains function (Loreau 2010). A
variety of theoretical models all predict a negative relationship
between species richness and ecosystem variability (Lehman and Tilman
2000, Ives and Hughes 2002, Loreau and {{de Mazancourt}} 2013), and
experimental tests tend to support such a prediction (Tilman et al.
2006, Hector et al. 2010).

However, the ability of biodiversity-ecosystem functioning (BEF)
experiments to accurately represent real-world dynamics is debated
(Wardle 2016, Eisenhauer et al. 2016). Much of the debate centers around
the fact that BEF experimental protocols do not allow species gains to
offset species losses. Theoretical work on diversity-variability
relationships typically suffers from the same limitation: it recognizes
the role of environmental variability in driving population fluctuations
which destabilize ecosystems, but ignores the potential for
environmental variability to promote species richness and thereby help
stabilize ecosystems (Loreau 2010, but see Chesson et al. 2001).

Fluctuating environmental conditions are an important ingredient for
stable species coexistence, both in theoretical models (Chesson 2000a,
Chesson et al. 2004) and in natural communities (C{á}ceres 1997,
Descamps-Julien and Gonzalez 2005, Adler et al. 2006, Angert et al.
2009). Such ``fluctuation-dependent'' coexistence requires that species
have unique environmental responses and that environmental conditions
vary so that each species in a community experiences favorable and
unfavorable conditions, which prevents competitive exclusion (Chesson
2000a). When coexistence is maintained by a fluctuation-dependent
mechanism, an increase in environmental variability might lead to an
increase in species richness and, consequently, a decrease in ecosystem
variability. However, increasing environmental variability may also
increase ecosystem variability by increasing the fluctuations of
individual species, regardless of species richness. These countervailing
effects of environmental variability present an interesting paradox:
while increasing variability should increase ecosystem variability, that
decrease might be offset by an increase in richness. Such a paradox
complicates predictions about how ecosystems will respond to predicted
departures from historical ranges of environmental variability.

The opposing effects of environmental variability on ecosystem
variability might explain the mixed results from empirical studies on
the diversity-variability relationship. Observational tests of the
diversity-variability relationship, which require sampling across
natural diversity gradients, have yielded negative (Hautier et al.
2014), neutral (Valone and Hoffman 2003, Cusson et al. 2015), and
positive (Sasaki and Lauenroth 2011) relationships. In a meta-analysis
of diversity-variability relationships based on observational studies in
terrestrial ecosystems, Jiang and Pu (2009) found no significant
evidence for an effect of species richness on ecosystem variability.
\tc{blue}{This may be because environmental variability varies across natural diversity gradients, affecting both richness and ecosystem variability.}
The idiosyncratic results of these empirical studies contrast with the
consistent conclusions from theoretical work that ignores feedbacks
between variability and richness.

The gap between theoretical expectations and empirical results of
richness-variability relationships might reflect the divergence of
theory developed to explain species coexistence and theory developed to
explain diversity and ecosystem variability. One reason these two
disciplines have diverged is because they have focused on different
questions. Diversity-variability studies typically ask how ecosystem
variability responds to different levels of species richness at a given
level of environmental variability (reviewed in Kinzig et al. 2001,
Loreau 2010), whereas coexistence studies ask how species richness
responds to different levels of environmental variability (Chesson and
Warner 1981).

To reconcile these two perspectives, we extend theory on the
relationship between species richness and ecosystem variability to cases
in which species coexistence explicitly depends on environmental
fluctuations and species-specific responses to environmental conditions.
We focus on the temporal storage effect and relative nonlinearity using
a general consumer-resource model. We use the model to investigate two
questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Does the diversity-variability relationship remain negative when
  species coexistence is maintained by the temporal storage effect or
  relative nonlinearity?
\item
  How does increasing environmental variability impact ecosystem
  variability when coexistence depends on the storage effect or relative
  nonlinearity?
\end{enumerate}

@@

During a growing season, consumer-resource dynamics are modeled as two
differential equations: \vspace{-3em}

\begin{align}
\frac{\text{d}N_{i}}{\text{d}t} &= \epsilon_if_{i}(R)N_{i}, \quad t \ne \tau_k\\
\frac{\text{d}R}{\text{d}t} &= - \sum\limits_{i}f_{i}(R)N_{i}, \quad t \ne \tau_k
\end{align}\vspace{-3em}

@@

Along with resource uptake, consumer population growth depends on the
production of dormant biomass (\(D\)), the activation of dormant biomass
to live biomass (\(D \rightarrow N\)), and the survival of living
biomass from one year to the next. The biomass of each species' states
at the start of a growing season are equal to \vspace{-3em}

\begin{align}
  D_{i}(\tau_k^+) &= (1-\gamma_{i,\tau_k})[\alpha_i N_{i}(\tau_k) + D_{i}(\tau_k)](1-\eta_i) \\
  N_{i}(\tau_k^+) &= (1-\alpha_i)N_{i}(\tau_k) + \gamma_{i,t}[\alpha_i N_{i}(\tau_k) + D_{i}(\tau_k)] (1-\eta_i),
\end{align}\vspace{-3em}

@@

For the storage effect to operate, we need species-specific responses to
environmental variability, density-dependent covariance between
environmental conditions and competition (\emph{EC} covariance), and
subadditive population growth (Chesson 1994, 2000b). Long-term
coexistence is possible when all species can increase when rare. In the
storage effect, rare species increase by escaping the effects of
\emph{EC} covariance. This happens because common species will
experience greater than average competition (\emph{C}) in good
environment (\emph{E}) years because common species cannot avoid
intraspecific competition. Rare species can escape intraspecific
competition and have the potential to increase rapidly in a good
\emph{E} year. \emph{EC} covariance emerges in our model because
dormant-to-live transition rates (\(\gamma\)) are species-specific and
vary through time. In a high \(\gamma\) year for a common species,
resource uptake will be above average, while in a high \(\gamma\) year
for a rare species, resource uptake will be below average, as long as
\(\gamma\)s are not perfectly correlated between species. Subadditive
population growth buffers populations against large population decreases
in unfavorable years. It is included in our model through a dormant
stage with very low death rates, which limits large population declines
in bad \emph{E} years.

We generated sequences of (un)correlated dormant-to-live state
transition rates (\(\gamma\)) for each species by drawing from
multivariate normal distributions with mean 0 and a variance-covariance
matrix (\(\Sigma(\gamma)\)) of \vspace{-3em}

\begin{align}
\Sigma(\gamma) = 
\begin{bmatrix}
1 & \rho_{1,2} & \rho_{1,3} & \rho_{1,4} \\
\rho_{2,1} & 1 & \rho_{2,3} & \rho_{2,4} \\
\rho_{3,1} & \rho_{3,2} & 1  & \rho_{3,4} \\
\rho_{4,1} & \rho_{4,2} & \rho_{4,3} & 1  \\
\end{bmatrix}
\sigma_{E}^2,
\end{align}\vspace{-2em}

\noindent where \(\sigma^2_{E}\) is the variance of the environmental
cue and \(\rho_{i,j}\) is the correlation between species \emph{i}'s and
species \emph{j}'s transition rates. \(\rho\) must be less than 1 for
stable coexistence, and in all simulations constrained all
\(\rho_{i,j}\)'s to be equal. In a two-species community, the inferior
competitor has the strongest potential to persist when \(\rho=-1\)
(perfectly uncorrelated transition rates). However, in a four-species
community the minimum possible correlation among species is -1/3 given
our constraints that all \(\rho\)'s are equal and that
\(\Sigma(\gamma)\) must be positive-definite. We used the \texttt{R}
function \texttt{mvrnorm} to generate sequences of (un)correlated
variates \emph{E} that we converted to germination rates in the 0-1
range: \(\gamma = e^E / \left(1 + e^E \right)\).

@@

For example, in Fig. 1C we show two species' resource uptake curves that
are relatively nonlinear. Species B has the lowest \(R^*\) and would
competitively exclude species A in the absence of environmental
fluctuations. But, fluctuating resource supplies can benefit species A
because it can take advantage of relatively high resource levels due its
higher saturation point. Stable coexistence is only possible, however,
if each species creates a disadvantage for itself when abundant. This
occurs in our model because when a resource conservative species (e.g.,
species B in Fig. 1C) is abundant, resource levels will remain high for
a longer period of time after a pulse because its draw down of resources
saturates. Likewise, when a resource acquisitive species (e.g., species
A in Fig. 1C) is abundant, it quickly draws down resources to levels
that favor resource conservative species. Such reciprocity ensures
species can increase when rare and stabilizes coexistence (Armstrong and
McGehee 1980, Chesson 2000a, Chesson et al. 2004).

@@

To explore how fluctuation-dependent coexistence can affect the
diversity-variability relationship, we simulated the model with four
species under two scenarios for each coexistence mechanism. First, we
allowed the variance of the environment to determine how many species
can coexist, akin to a community assembly experiment with a species pool
of four species. To do this, we simulated communities with all species
initially present across a gradient of annual resource variability (for
relative nonlinearity) or environmental cue variability (for the storage
effect). Second, we chose parameter values that allowed coexistence of
all four species and then performed species removals, akin to a
biodiversity-ecosystem function experiment. The two simulation
experiments correspond to (i) sampling ecosystem function across a
natural gradient of species richness and (ii) sampling ecosystem
function across diversity treatments within a site.

To understand how increasing environmental variability will impact
ecosystem variability when coexistence is fluctuation-dependent, we
simulated the model over a range of species pool sizes and environmental
cue or resource variability. For each size of species pool (1, 2, 3, or
4 species), we simulated the model at 15 evenly-spaced levels of
environmental cue (range = 0.1,2) for the storage effect and 25
evenly-spaced levels of resource variability (range = 0.1,1.4) for
relative nonlinearity. We also explored the influence of asymmetries in
species' competitive abilities and correlations in species'
environmental responses within the storage effect model.

Under relative nonlinearity, species' resource response curves (Fig.
S1-1) reflect traits that determine the temporal variability of each
species' population growth. ``Stable'' species achieve maximum resource
uptake at low resource levels, but their maximum uptake rates are
modest. For these species, population responses to resource fluctuations
are buffered. ``Unstable'' species have very high maximum uptake rates,
which they only achieve when resource availability is high, leading to
large population fluctuations. The difference in the intrinsic stability
of these two kinds of species makes our simulations sensitive to initial
conditions. Therefore, we ran two sets of simulations for relative
nonlinearity: beginning with either stable or unstable species as a
reference point For example, if species A is the most stable species and
species D is the least stable, we ran simulations where A then B then C
then D were allowed to invade. We then ran simulations with that order
reversed.

All simulations were run for 5,000 growing seasons of 100 days each. We
averaged biomass over the growing season, and those yearly values were
used to calculate total community biomass in each year. After discarding
an initial 500 seasons to reduce transient effects on our results, we
calculated the coefficient of variation (\emph{CV}) of summed species
biomass through time, which represents ecosystem variability, the
inverse of ecosystem stability. We calculated species richness as the
number of species whose average biomass was greater than 1 over the
course of the simulation. Parameter values for specific results are
given in figure captions. Within-season dynamics were solved given
initial conditions using the package \texttt{deSolve} (Soetaert et al.
2010) in \texttt{R} (Team 2013). \texttt{R} code for our model function
is in the Supporting Information Appendix S1. All model code has been
deposited on Figshare (\emph{link}) and is available on GitHub at
\url{http://github.com/atredennick/Coexistence-Stability}.

@@

When we allowed the variance of the environment to determine which of
four initial species coexisted, similar to a community assembly
experiment, we found a positive relationship between richness and
variability (Fig. 2A,C). This was true for the storage effect, where
coexistence is maintained by fluctuating live-to-dormant transition
rates (\(\gamma\)), and for relative nonlinearity, where coexistence is
maintained by annual resource pulses. The relationship is driven by the
fact that increasing environmental variability increases the strength of
both coexistence mechanisms (Fig. S1-2). More variable conditions
promoted species richness, creating a positive relationship between
diversity and variability.

When we performed species removals but held environmental variability at
a level that allows coexistence of all four species, similar to a
biodiversity-ecosystem functioning experiment, we found a negative
diversity-variability relationship (Fig. 2B,D). Scatter around the
relationship was small for the storage effect because all species has
similar temporal variances. Regardless of species identity, the presence
of more species always stabilizes ecosystem functioning through
portfolio effects. In contrast, scatter around the relationship was
larger for relative nonlinearity (Fig. 2D) because species with
different resource uptake curves have different population variances.
Depending on which species were present, two-species communities were
sometimes less variable than three-species communities.

To understand how much species additions can stabilize ecosystem
functioning as environmental variability increases, we simulated our
model over a range of environmental variance and species pool sizes.
Realized species richness increased with environmental variability and,
in some cases, increases in richness completely offset the effect of
increasing environmental variability on ecosystem variability. More
species rich communities were less variable on average and increased in
ecosystem \emph{CV} at a slower rate than communities with fewer species
(e.g., lower slopes in log-log space; Fig. S1-3).

The dampening effect of fluctuation-dependent coexistence on increasing
environmental variability depends on the specific traits (parameter
values) of the species in the regional pool. Moderate competition makes
it more difficult for new species to enter the local community, but once
they do enter, variability is similar between communities with low and
moderate competition (Fig. 3; compare top and bottom panels). Moderate
competition does increase the rate at which ecosystem \emph{CV}
increases with environmental variance (Fig. S1-3) because the abundance
of inferior competitors is reduced and they do not influence ecosystem
\emph{CV} as much as when competition is low. The correlation of
species' environmental responses also mediates the relationship between
environmental variance, species richness, and ecosystem \emph{CV}: lower
correlations make it easier for new species to enter the community and
contribute to porfolio effects (Fig. 3).

In communities where species coexist via relative nonlinearity, the
extent to which species additions buffer ecosystem stability against
increases in environmental variability depends on the species traits of
immigrating species and the order in which they enter the community.
When additional species, which immigrate from the regional pool, are
less intrinsically stable than the resident species, ecosystem
variability increases at a constant rate even as species are added (Fig.
4A). If more stable species colonize, species additions buffer the
ecosystem from increasing environmental variability (Fig. 4B). When only
the most unstable (resource acquisitive) species is present, ecosystem
variability initially declines as resource variability increases (Fig.
4B).
\tc{blue}{This occurs because community biomass is very low at low average resource levels for this species, which inflates the *CV*.}

@@

Theory developed for biodiversity-ecosystem function experiments
emphasizes that increases in species richness should reduce ecosystem
variability. Consistent with theoretical expectations from models and
with results from biodiversity-ecosystem functioning experiments, in
which species coexistence is maintained by fluctuation-independent
mechanisms, we produced a negative diversity-variability relationship
(Fig. 2B,D) from a model with fluctuation-dependent species coexistence
(also see Chesson et al. 2001). This agreement is encouraging because
species almost certainly coexist by some combination of
fluctuation-independent (e.g., resource partitioning) and
fluctuation-dependent mechanisms. By extending theory to communities
where species richness is explicitly maintained by temporal variability,
we have gained confidence that experimental findings are generalizable
to many communities. In local settings where environmental variability
is relatively homogeneous, reductions in the number of species should
increase the variability of ecosystem functioning, regardless of how
coexistence is maintained.

When we allowed communities to assemble at sites across a gradient of
environmental variability we discovered a positive relationship between
species richness and ecosystem variability (Fig. 2A,C). While surprising
when viewed through the lens of biodiversity-ecosystem functioning
theory and experimental findings, such a relationship is predicted by
theory on coexistence in fluctuating environments. Environmental
variability is a prerequisite for the storage effect and relative
nonlinearity to stabilize coexistence (Chesson 2000a). These mechanisms
can translate increased variability into higher species richness (Fig.
S1-2), but the increase in environmental variability also increases
ecosystem variability.

Our results may explain why deviations from the negative
diversity-variability relationship often come from observational studies
(Jiang and Pu 2009). Observational studies must rely on natural
diversity gradients, which do not control for differences in
environmental variability among sites. If species richness depends on
environmental variability, it is entirely possible to observe positive
diversity-variability relationships. For example, Sasaki and Lauenroth
(2011) found a positive relationship between species richness and the
temporal variability of plant abundance in a semi-arid grassland. Their
data came from a six sites that were 6 km apart. While Sasaki and
Lauenroth explained their results in terms of dominant species' effects,
it is also possible that each site experienced sufficiently different
levels of environmental variability to influence species coexistence.
DeClerck et al. (2006) also found a positive diversity-variability when
sampling conifer richness and the variability of productivity across a
large spatial gradient in the Sierra Nevada.

While our modeling results show that fluctuation-dependent coexistence
can create positive diversity-variability relationships, whether such
trends are detected will depend on the particular traits of the species
in the community and the relative influence of fluctuation-dependent and
fluctuation-independent coexistence mechanisms. Thus, our results may
also help explain observational studies where no relationship between
diversity and variability is detected. For example, Cusson et al. (2015)
found no relationship between species richness and variability of
abundances in several marine macro-benthic ecosystems. Many of their
focal ecosystems were from highly variable intertidal environments. If
coexistence was at least in part determined by environmental
fluctuations, then the confounding effect of variability and species
richness could compensate any direct effect of species richness on
variability. Previous theoretical work showed how environmental
variation can mask the effect of species diversity on ecosystem
productivity when sampling across sites (Loreau 1998). Our mechanistic
model extends that conclusion to ecosystem variability.

Whether coexistence is fluctuation-independent or fluctuation-dependent
becomes especially important when we consider how ecosystem variability
responds to increasing environmental variability. In the
fluctuation-independent case, species richness is essentially fixed
because the species' inequalities that determine coexistence (niche and
fitness differences) are not linked to environmental variability.
Therefore, increasing environmental variability will always increase
ecosystem variability by increasing the fluctuations of individual
species' abundances.
